---
title: "Mutually Exclusive Phenotypes"
author: "Amandeep Rathee"
date: '`r format(Sys.Date(), "%B-%d-%Y")`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rphenoscape)
```


This vignette showcases a function `are_mutually_exclusive` that establishes whether two phenotypes are mutually exclusive based on the phenoscape KB data.

```{r include=TRUE, echo=TRUE, warning=FALSE, render=FALSE}

are_mutually_exclusive <- function(phenotype.a, phenotype.b, studies, verbose=FALSE){
    
    # mutually_exclusive takes three values:
    # -1: default; we assume that before checking mutual exclusion, phenotypes are neither mutually exclusive nor mutually compatible
    #  0: data suggests mutual compatibility
    #  1: data suggests mutual exclusion
    mutually_exclusive <- -1
    
    # get charstates dataframe
    character_states <- charstates(c(phenotype.a, phenotype.b))
    
    # filter states based on phenotypes and studies
    states.a <- character_states$state.id[(character_states$phenotype.id == phenotype.a) & (character_states$study.id %in% studies)]
    states.b <- character_states$state.id[(character_states$phenotype.id == phenotype.b) & (character_states$study.id %in% studies)]
    
    
    # determine mutual exclusivity only if both study vectors are non-empty
    if((length(states.a) > 0) & (length(states.b) > 0)){ 
        
        # get character vectors
        chars.a <- unique(character_states$character.id[character_states$phenotype.id == phenotype.a])
        chars.b <- unique(character_states$character.id[character_states$phenotype.id == phenotype.b])
        
        # check intersection of character vectors
        char_matches <- sum(chars.a %in% chars.b)
        
        if(char_matches == 0){ 
            # CASE 1: if no characters match --> phenotypes are mutually compatible
            if(verbose){cat("No character match.\n")}
            mutually_exclusive <- 0
            
        } else { 
            # CASE 2: if one or more characters match --> investigate further
            
            # compute intersection of the character and state vectors
            char_state_match_count <- get_char_state_match_count(character_states, phenotype.a, phenotype.a)
            
            # number of common characters between phenotype A and B
            char_match_count <- char_state_match_count[1]
            
            # number of states that are different among the common characters between phenotype and B
            state_mismatch_count <- char_state_match_count[2] 
            
            if(char_match_count == state_mismatch_count){
                # CASE 2-I: for matched characters, if no states are common among the matched characters --> phenotypes are mutually exclusive
                if(verbose){cat("Character match but no state match.\n")}
                mutually_exclusive <- 1
            } else {
                # CASE 2-II: if, for any matched characters, any two corresponding states match --> phenotypes are mutually compatible
                if(verbose){cat("Character match and state match.\n")}
                mutually_exclusive <- 0
            }
        }
        
        if(mutually_exclusive == -1){warning('Insufficient evidence to determine mutual exclusivity.')}  
        mutually_exclusive
        
    } else {
        if(length(studies.a) == 0){
            stop("Phenotype A is not part of any of the studies passed. Both phenotypes are required to be part of at least one study.")
        } else if (length(studies.b) == 0){
            stop("Phenotype B is not part of any of the studies passed. Both phenotypes are required to be part of at least one study.")
        } else {
            stop("Phenotypes A and B are not part of any of the studies passed. Both phenotypes are required to be part of at least one study.")
        }
    }
}

get_char_state_match_count <- function(char_state_dataframe, phenotype.a, phenotype.b){
    
    # filter phenotypes
    char_state_df_a <- char_state_dataframe[char_state_dataframe$phenotype.id == phenotype.a, ]
    char_state_df_b <- char_state_dataframe[char_state_dataframe$phenotype.id == phenotype.b, ]
    
    char_match_count     <- 0   # number of common characters between phenotype A and B.
    state_mismatch_count <- 0   # number of states that are different among the common characters between phenotype and B.
    
    for(index.a in 1:nrow(char_state_df_a)){          # iterate over rows of dataframe that contains phenotype A
        for (index.b in 1:nrow(char_state_df_b)){     # iterate over rows of dataframe that contains phenotype B
            
            if(char_state_df_a[index.a, 'character.id'] == char_state_df_b[index.b, 'character.id']){
                char_match_count = char_match_count + 1
                if( char_state_df_a[index.a, 'state.id'] != char_state_df_b[index.b, 'state.id'] ){
                    state_mismatch_count = state_mismatch_count + 1
                }
            }
        }
    }
    
    c(char_match_count, state_mismatch_count)
}

```

# Examples of mutual compatibility.

### Case 1: When there are no common characters between phenotypes A and B.

#### Test case 1

```{r}
# get list of phenotypes
phenotypes <- get_phenotypes(entity = "hyomandibular bone", taxon = "Hypancistrus")

# get list of studies
studies <- pk_get_study_list(entity = "hyomandibular bone", taxon = "Hypancistrus")

# check mutual exclusivity between any two phenotypes from the phenotype list
phenotype.a <- phenotypes$id[phenotypes$label == 'hyomandibular bone in contact with prootic bone']
phenotype.b <- phenotypes$id[phenotypes$label == 'hyomandibular bone in contact with quadrate bone']
study <- studies$id[studies$label == 'Armbruster (2004)']
are_mutually_exclusive(phenotype.a, phenotype.b, study, verbose=TRUE)
```



#### Test case 2: 

```{r}
# get list of phenotypes
phenotypes <- get_phenotypes(entity = "hyomandibular bone", taxon = "Hypancistrus")

# get list of studies
studies <- pk_get_study_list(entity = "hyomandibular bone", taxon = "Hypancistrus")

# check mutual exclusivity between any two phenotypes from the phenotype list
phenotype.a <- phenotypes$id[phenotypes$label == 'hyomandibular bone in contact with prootic bone']
phenotype.b <- phenotypes$id[phenotypes$label == 'pterotic-posttemporal-supracleithrum in contact with hyomandibular bone']
study <- studies$id[studies$label == 'Armbruster (2004)']
are_mutually_exclusive(phenotype.a, phenotype.b, study, verbose=TRUE)

```

### Case 2: When there are common states among characters between phenotypes A and B.

#### Test case 3

```{r}
# get list of phenotypes
phenotypes <- get_phenotypes(taxon = 'Abramites hypselonotus')

# get list of studies
studies <- pk_get_study_list(taxon = 'Abramites hypselonotus')

# check mutual exclusivity between any two phenotypes from the phenotype list
phenotype.a <- phenotypes$id[phenotypes$label == 'adductor mandibulae structure preopercle horizontal limb']
phenotype.b <- phenotypes$id[phenotypes$label == 'adductor mandibulae structure preopercle horizontal limb']
study <- studies$id[studies$label == 'Dillman et al. (2016)']
are_mutually_exclusive(phenotype.a, phenotype.b, study, verbose=TRUE)

```

#### Test case 4

```{r}
# get list of phenotypes
phenotypes <- get_phenotypes(taxon = 'Abramites hypselonotus')

# get list of studies
studies <- pk_get_study_list(taxon = 'Abramites hypselonotus')

# check mutual exclusivity between any two phenotypes from the phenotype list
phenotype.a <- phenotypes$id[phenotypes$label == 'ventral region and (part_of some symplectic) decreased width']
phenotype.b <- phenotypes$id[phenotypes$label == 'ventral region and (part_of some symplectic) laminar']
study <- studies$id[studies$label == 'Dillman et al. (2016)']
are_mutually_exclusive(phenotype.a, phenotype.b, study, verbose=TRUE)
```



## Test mutual exclusivity among more than two phenotypes

```{r}
mutual_exclusivity_matrix <- function(phenotypes, studies){
    # return square matrix of size length(phenotypes)
    # matrix has the following values
    # -2: phenotypes present in the row and column are not part of the studies provided
    # -1: default; insufficient data to conclude mutual exclusivity or compatibility
    #  0: phenotypes belonging to row and column are mutually compatible
    #  1: phenotypes belonging to row and column are mutually exclusive
    
    # initialize matrix
    num_phenotypes <- length(phenotypes)
    mutual_exclusivity_matrix <- matrix(-1, nrow=num_phenotypes, ncol=num_phenotypes)
    
    # fill each cell of the matrix
    for (i in 1:num_phenotypes) {
        for (j in i:num_phenotypes) {
            
            if(i != j){ # compute mutual exclusivity among distinct phenotypes only
                mutual_exclusivity <- are_mutually_exclusive(phenotypes[i], phenotypes[j], studies)
                mutual_exclusivity_matrix[i, j] <- mutual_exclusivity
                mutual_exclusivity_matrix[j, i] <- mutual_exclusivity
            } else {    # phenotype is mutually compatible with itself
                mutual_exclusivity_matrix[i, j] <- 0
            }
        }
    }
    
    # return matrix
    mutual_exclusivity_matrix
}
```


```{r}
# get phenotypes
phenotypes <- get_phenotypes(entity = "hyomandibular bone", taxon = "Hypancistrus")
phenotypes <- c(phenotypes$id[phenotypes$label == 'hyomandibular bone in contact with prootic bone'],
                phenotypes$id[phenotypes$label == 'hyomandibular bone in contact with quadrate bone'],
                phenotypes$id[phenotypes$label == 'hyomandibular bone position opercle'],
                phenotypes$id[phenotypes$label == 'hyomandibular bone structure metapterygoid'])
# get studies
studies <- pk_get_study_list(entity = "hyomandibular bone", taxon = "Hypancistrus")
study <- studies$id[studies$label == 'Armbruster (2004)']

# check mutual exclusivity
mutual_exclusivity_matrix(phenotypes, study)

```



TODOS:

1. Examples of mutual exclusivity
2. Test multiple studies
